---
title: "Lecture 01"
subtitle: "Introduction to Environmental Economics"
author: Ivan Rudik
date: AEM 4510
output:
  xaringan::moon_reader:
    css: ['default', 'metropolis', 'metropolis-fonts', 'my-css.css']
    # self_contained: true
    nature:
      highlightStyle: github
      highlightLines: true
      countIncrementalSlides: false
      ratio: '16:9'

---
exclude: true
```{r setup}
if (!require("pacman")) install.packages("pacman")
pacman::p_load(
  tidyverse, xaringanExtra, rlang, patchwork, nycflights13
)
options(htmltools.dir.version = FALSE)
knitr::opts_hooks$set(fig.callout = function(options) {
  if (options$fig.callout) {
    options$echo <- FALSE
  }
knitr::opts_chunk$set(echo = TRUE, fig.align="center")
  options
})
```
```{r xaringanExtra, echo = FALSE}
xaringanExtra::use_xaringan_extra(c("tile_view", "panelset", "webcam"))

```
```{r echo=FALSE}
xaringanExtra::style_panelset(panel_tab_color_active = "red")
```

---

# Roadmap

- What is environmental economics?
- What are the goals for this class?
- Microeconomics recap

---

class: inverse, center, middle
name: what_is_enviro

# What is environmental economics?

<html><div style='float:left'></div><hr color='#EB811B' size=1px width=796px></html>


---

# What's enviro econ?

First, what is economics?

--

.hi[Economics:] the study of how agents (people, firms, etc) make choices with scarce resources and the social results of these choices

--

Everything is scarce compared to wants and needs

--

We need to choose among alternatives and make trade offs

--

These ideas can be applied to the environment

---

# What is enviro econ?

.hi[Environmental Economics:] the application of economics to the study of the environment as a resource or good

--

Environmental economics helps us understand things like:

--

- The value of mitigating pollution

--

How agents will response to climate change policies

--

Whether investment tax credits for wind power are cost-effective

---

class: inverse, center, middle
name: goals

# What are the goals of this class?

<html><div style='float:left'></div><hr color='#EB811B' size=1px width=796px></html>


---

# What are the goals of this class?

1. Enhance your understanding of	how economists think about solving real environmental problems
--

2. Prepare you for after college:
  - Industry jobs (energy, transportation, finance)
  - Public sector and NGOs (EPA, DOE, RFF, Brookings, Federal Reserve)
  - Graduate programs
  
--

3. Learn both the theory and .hi[applications] of environmental economics

---

class: inverse, center, middle
name: micro_recap

# Microeconomics recap

<html><div style='float:left'></div><hr color='#EB811B' size=1px width=796px></html>


---

# Micro recap