class: inverse, center, middle
name: solar

# Solar tariffs

<html><div style='float:left'></div><hr color='#EB811B' size=1px width=796px></html>

---

# Solar tariffs: the opposite-sign case

```{r fig-solar-mirror-case, out.width = "100%", fig.width = 11.2, fig.height = 3.9, echo = FALSE, fig.align = "center"}
st_plot_mirror_case("both")
```

- Carbon tariffs tax a .hi[dirty good]
- Solar tariffs tax a .hi[clean input]
- The missing wedge is on the .hi[benefit side], not the cost side

---

# Free trade benchmark for solar modules

.pull-left[
```{r fig-solar-free-trade, out.width = "100%", fig.width = 5.6, fig.height = 4.3, echo = FALSE, fig.align = "center"}
st_plot_solar_tariff("free_trade")
```
]

.pull-right[

At price $P_W$:
- Domestic manufacturing is $q_i^S$
- Domestic purchases/installations are $q_i^D$
- Imports are $q_i^D - q_i^S$

]

---
count: false
# The tariff wedge for solar modules

.pull-left[
```{r fig-solar-tariff-price, out.width = "100%", fig.width = 5.6, fig.height = 4.3, echo = FALSE, fig.align = "center"}
st_plot_solar_tariff("price")
```
]

.pull-right[

- A tariff raises the domestic module price
- Domestic manufacturers can charge the higher price
- Imported units now enter at the tariff-inclusive price $P_W + \tau$

]

---
count: false
# Quantity effects of a solar tariff

.pull-left[
```{r fig-solar-tariff-quantities, out.width = "100%", fig.width = 5.6, fig.height = 4.3, echo = FALSE, fig.align = "center"}
st_plot_solar_tariff("quantities")
```
]

.pull-right[

- Domestic manufacturing rises to $q_{i\tau}^S$
- Domestic purchases/installations fall to $q_{i\tau}^D$
- Imports shrink
- In the simple 2-country model, foreign production falls with imports

]

---
count: false
# Same private tariff geometry as before

.pull-left[
```{r fig-solar-tariff-welfare, out.width = "100%", fig.width = 5.6, fig.height = 4.3, echo = FALSE, fig.align = "center"}
st_plot_solar_tariff("welfare")
```
]

.pull-right[

- Producer gain = $a+b$
- Buyers/installers lose = $a+b+c+d+e$
- Government revenue = $d$
- Private deadweight loss = $c+e$
- .hi[Privately], a solar tariff looks just like any ordinary tariff

]

---
<!-- Sources: One Earth paper on climate and health benefits of imported solar PV in the US: https://www.sciencedirect.com/science/article/abs/pii/S2590332225002933 -->
# Solar deployment has a positive externality

.pull-left[
```{r fig-solar-externality-private, out.width = "100%", fig.width = 5.6, fig.height = 4.3, echo = FALSE, fig.align = "center"}
st_plot_solar_externality("private")
```
]

.pull-right[

- Solar deployment displaces fossil generation

]

---
count: false
# Solar deployment has a positive externality

.pull-left[
```{r fig-solar-externality-smb, out.width = "100%", fig.width = 5.6, fig.height = 4.3, echo = FALSE, fig.align = "center"}
st_plot_solar_externality("smb")
```
]

.pull-right[

- Solar deployment displaces fossil generation
- Private demand understates social value

]

---
count: false
# Solar deployment has a positive externality

.pull-left[
```{r fig-solar-externality-smb-final, out.width = "100%", fig.width = 5.6, fig.height = 4.3, echo = FALSE, fig.align = "center"}
st_plot_solar_externality("smb")
```
]

.pull-right[

- Solar deployment displaces fossil generation
- Private demand understates social value
- The efficient quantity is higher than the private-market quantity

]

---

# Solar tariff with deployment benefits

.pull-left[
```{r fig-solar-tariff-social, out.width = "100%", fig.width = 5.6, fig.height = 4.3, echo = FALSE, fig.align = "center"}
st_plot_solar_externality("tariff_social")
```
]

.pull-right[

- The standard tariff losses $c+e$ are still there
- There is an extra environmental loss $g$
- If domestic and foreign manufacturing have the same environmental intensity, then $\Delta W_i^{social} = -(c + e + g)$

]

---

# When could a solar tariff help the climate?

.pull-left[
```{r fig-solar-manufacturing-offset, out.width = "100%", fig.width = 5.6, fig.height = 4.3, echo = FALSE, fig.align = "center"}
st_plot_solar_externality("manufacturing_offset")
```
]

.pull-right[

- Cleaner domestic manufacturing can create an upstream environmental gain
- $\Delta W_i^{social} = h - (c + e + g)$
- A solar tariff helps the climate only if $h$ is large enough to offset $g$ and the usual tariff distortions
- If domestic manufacturing is .hi[dirtier], then $h < 0$ and the case gets weaker

]

---
<!-- Sources: NBER working paper on strategic avoidance and welfare impacts of U.S. solar tariffs: https://www.nber.org/system/files/working_papers/w34401/w34401.pdf -->
# Trade diversion in practice

- Country-specific tariffs often shift production to .hi[third countries], not just back home
- That weakens the domestic-manufacturing argument
- It also makes climate accounting messier
- This is the solar analogue of .hi[leakage / resource shuffling]
- Chinese manufacturers' offshore manufacturing share rose from `5%` in `2012` to `20%` by `2019`

<!-- TODO numerical examples: (i) linear D and S with constant deployment wedge B; (ii) cleaner domestic manufacturing case with h > 0; (iii) trade-diversion case with third-country supply. -->
