---
title: "Lecture 09"
subtitle: "Carbon Tariffs"
author: Ivan Rudik
date: AEM 4510
output:
  xaringan::moon_reader:
    css: ['default', 'metropolis', 'metropolis-fonts', 'my-css.css']
    # self_contained: true
    nature:
      highlightStyle: github
      highlightLines: true
      countIncrementalSlides: false
      ratio: '16:9'

<!-- Auto-rebuild: ls slides/09-slides-carbon-tariffs.Rmd | entr -c Rscript -e "rmarkdown::render('slides/09-slides-carbon-tariffs.Rmd')" -->

---
exclude: true
```{r setup}
if (!require("pacman")) install.packages("pacman")
pacman::p_load(
  tidyverse, xaringanExtra, rlang, patchwork
)
source("R/video_helpers.R")
source("R/09-carbon-tariffs-figures.R")
options(htmltools.dir.version = FALSE)
knitr::opts_hooks$set(fig.callout = function(options) {
  if (options$fig.callout) {
    options$echo <- FALSE
  }
knitr::opts_chunk$set(echo = TRUE, fig.align="center")
  options
})
```
```{r xaringanExtra, echo = FALSE}
xaringanExtra::use_xaringan_extra(c("tile_view", "panelset", "webcam"))

```
```{r echo=FALSE}
xaringanExtra::style_panelset(panel_tab_color_active = "red")
```

---

# Roadmap

1. What are carbon tariffs?
2. Basic trade theory: supply, demand, and tariffs
3. Adding pollution externalities to trade
4. Unilateral carbon taxes and leakage
5. Carbon tariffs to prevent leakage
6. Carbon tariffs alone
7. World carbon taxes
8. Different pollution intensities across countries

---

class: inverse, center, middle
name: overview

# What are carbon tariffs?

<html><div style='float:left'></div><hr color='#EB811B' size=1px width=796px></html>

---

# Carbon tariffs

Carbon tariffs are tariffs imposed on imports based on their .hi[carbon content]

--

Why would we want carbon tariffs?

---

# Understanding border carbon adjustments

```{r, echo = FALSE}
local_video("videos/09-carbon-tariffs-bca-explained.mp4")
```

---

# Carbon tariffs

Why would we want carbon tariffs?

--

1. .hi[Leakage:] if country $i$ imposes a carbon tax, production may shift to country $j$ with no carbon tax

--

2. .hi[Competitiveness:] domestic firms face higher costs than foreign competitors

--

Carbon tariffs aim to "level the playing field" and reduce leakage

---

# Carbon tariffs in practice

.hi[EU Carbon Border Adjustment Mechanism (CBAM):]
- Phasing in 2023-2026, full implementation 2026
- Covers cement, iron/steel, aluminum, fertilizers, electricity, hydrogen
- Importers must purchase certificates matching EU ETS price

--

.hi[UK CBAM:] announced for 2027

--

.hi[US proposals:] various bills proposed but none passed yet

---

# This lecture

We will use a simple partial equilibrium framework from Brunel and Levinson (2024) to understand:

--

1. How carbon tariffs work in theory

--

2. When they improve welfare

--

3. What complications arise in practice

--

The key insight: carbon tariffs are equivalent to a .hi[consumption tax] plus a .hi[production subsidy]

---

class: inverse, center, middle
name: trade

# Basic trade theory

<html><div style='float:left'></div><hr color='#EB811B' size=1px width=796px></html>

---

# A simple trade model

Consider a small open economy (country $i$) that:
- Takes world prices as given
- Can import or export a homogeneous good

--

Key curves:
- $S$: domestic supply
- $D$: domestic demand
- $p_w$: world price

---

# Free trade equilibrium

<!-- FIGURE 1: Standard tariff diagram showing S, D curves, world price, imports, and deadweight loss triangles a, b from tariff -->

```{r fig1-tariff, out.width = "900px", fig.width = 11, fig.height = 4.25, echo = FALSE, fig.align = "center"}
ct_plot_figure_1(stage = "free_trade")
```

With free trade at price $p_w$:
- Domestic production: $q_i^S$
- Domestic consumption and imports: $q_i^D$ and $q_i^D - q_i^S$

---
count: false
# Adding a tariff

```{r fig1-tariff-stage2, out.width = "900px", fig.width = 11, fig.height = 4.25, echo = FALSE, fig.align = "center"}
ct_plot_figure_1(stage = "tariff")
```

A tariff $\tau$ raises the domestic price to $p_w + \tau$:
- Production rises to $q_{i\tau}^S$ and consumption falls to $q_{i\tau}^D$
- Imports fall to $q_{i\tau}^D - q_{i\tau}^S$

---
count: false
# Tariff revenue

```{r fig1-tariff-stage3, out.width = "900px", fig.width = 11, fig.height = 4.25, echo = FALSE, fig.align = "center"}
ct_plot_figure_1(stage = "welfare")
```

The tariff creates:
- Revenue: $R(\tau) = \tau (q_{i\tau}^D - q_{i\tau}^S)$
- Deadweight loss: triangles $a + b$

---
count: false
# Deadweight loss

```{r fig1-tariff-stage4, out.width = "900px", fig.width = 11, fig.height = 4.25, echo = FALSE, fig.align = "center"}
ct_plot_figure_1(stage = "welfare")
```

Triangle $a$ is a production distortion:
- The tariff shifts production from lower-cost foreign supply to higher-cost domestic production.

---
count: false
# Deadweight loss

```{r fig1-tariff-stage5, out.width = "900px", fig.width = 11, fig.height = 4.25, echo = FALSE, fig.align = "center"}
ct_plot_figure_1(stage = "welfare")
```

Triangle $b$ is a consumption distortion:
- The tariff eliminates units consumers value above the world price but below the tariff-inclusive price.

---

class: inverse, center, middle
name: pollution

# Adding pollution

<html><div style='float:left'></div><hr color='#EB811B' size=1px width=796px></html>

---

# Production creates pollution

Now suppose production of this good creates pollution

--

Let $e$ = emissions per unit of output (same everywhere for now)

--

Let $d$ = marginal damage from emissions (global pollutant)

--

The .hi[marginal social cost (MSC)] of production is: $MSC = S + e \times d$

--

MSC lies above the private supply curve by the external cost $e \times d$

---

# Free trade with pollution

<!-- FIGURE 2a: Country i with MSC curve above S, showing pollution externality -->

.pull-left[
```{r fig2a-pollution, out.width = "100%", fig.width = 5.4, fig.height = 4.3, echo = FALSE, fig.align = "center"}
ct_plot_figure_2_country(stage = "baseline")
```
]

.pull-right[

Free trade output is pinned down by domestic demand and supply at the world price.

]

---
count: false
# Free trade with pollution

.pull-left[
```{r fig2a-pollution-stage2, out.width = "100%", fig.width = 5.4, fig.height = 4.3, echo = FALSE, fig.align = "center"}
ct_plot_figure_2_country(stage = "msc")
```
]

.pull-right[

Country $i$'s MSC curve lies above its private supply curve.

The vertical gap is the external cost: $e \times d$.

]

---
count: false
# Free trade with pollution

.pull-left[
```{r fig2a-pollution-stage3, out.width = "100%", fig.width = 5.4, fig.height = 4.3, echo = FALSE, fig.align = "center"}
ct_plot_figure_2_country(stage = "damage")
```
]

.pull-right[

The green region is the domestic deadweight loss from producing beyond the efficient output level.

]

---
count: false
# Free trade with pollution

.pull-left[
```{r fig2a-pollution-stage4, out.width = "100%", fig.width = 5.4, fig.height = 4.3, echo = FALSE, fig.align = "center"}
ct_plot_figure_2_country(stage = "damage")
```
]

.pull-right[

At the world price $p_w$, marginal benefit is pinned down by world demand, while marginal social cost is given by $MSC$.

That triangle is the set of units for which marginal benefit exceeds private supply cost but is below marginal social cost.

]

---

# Global perspective

<!-- FIGURE 2b: World supply and demand with MSC -->

.pull-left[
```{r fig2b-world, out.width = "100%", fig.width = 5.4, fig.height = 4.3, echo = FALSE, fig.align = "center"}
ct_plot_figure_2_world()
```
]

.pull-right[

From a global perspective:
- Free-trade output is where $D = S$.
- Efficient output is where $D = MSC$.
- The green region is the global welfare gain from cutting output from $Q_W$ to $Q_W^*$.

]

---

# The externality problem

With a global pollutant like $CO_2$:

--

- Damages occur regardless of where emissions happen
- No country has incentive to fully internalize the externality
- Free trade leads to .hi[too much production] globally

--

A global carbon tax $= e \times d$ would achieve efficiency

--

But what if only .hi[one country] acts?

---

class: inverse, center, middle
name: unilateral

# Unilateral carbon tax

<html><div style='float:left'></div><hr color='#EB811B' size=1px width=796px></html>

---

# Country $i$ imposes a carbon tax

Suppose country $i$ imposes a carbon tax $\tau = e \times d$ on domestic production

--

This shifts country $i$'s supply curve up by $\tau$

--

But the .hi[world price doesn't change] (small open economy)

---

# Effects of unilateral carbon tax

<!-- FIGURE 3: Unilateral carbon tax showing tax revenue Rt, deadweight loss d, and leakage -->

.pull-left[
```{r fig3-unilateral, out.width = "100%", fig.width = 5.8, fig.height = 4.3, echo = FALSE, fig.align = "center"}
ct_plot_figure_3(stage = "baseline")
```
]

.pull-right[

Before the tax, the domestic market is at the free-trade benchmark with price $p_w$.

]

---
count: false
# Effects of unilateral carbon tax

.pull-left[
```{r fig3-unilateral-stage2, out.width = "100%", fig.width = 5.8, fig.height = 4.3, echo = FALSE, fig.align = "center"}
ct_plot_figure_3(stage = "tax")
```
]

.pull-right[

After the tax:
- Producers receive the net-of-tax price $p_w - t$
- Because $t$ equals the external damage, domestic output falls to the Figure 2 optimum $q_{it}^S$
- Consumers still face the world price $p_w$

]

---
count: false
# Effects of unilateral carbon tax

.pull-left[
```{r fig3-unilateral-stage3, out.width = "100%", fig.width = 5.8, fig.height = 4.3, echo = FALSE, fig.align = "center"}
ct_plot_figure_3(stage = "tax")
```
]

.pull-right[

Leakage comes next:
- Consumers in country $i$ still demand $q_i^D$ at price $p_w$
- Imports rise enough to offset the fall in domestic output
- Foreign production replaces home production: that is leakage

]

---
count: false
# Effects of unilateral carbon tax

.pull-left[
```{r fig3-unilateral-stage4, out.width = "100%", fig.width = 5.8, fig.height = 4.3, echo = FALSE, fig.align = "center"}
ct_plot_figure_3(stage = "welfare")
```
]

.pull-right[

From country $i$'s perspective:
- Consumers are no better or worse off: they still pay $p_w$ and consume $q_i^D$
- The government gains revenue $R_t$

]

---
count: false
# Effects of unilateral carbon tax

.pull-left[
```{r fig3-unilateral-stage5, out.width = "100%", fig.width = 5.8, fig.height = 4.3, echo = FALSE, fig.align = "center"}
ct_plot_figure_3(stage = "welfare")
```
]

.pull-right[

From country $i$'s perspective:
- Consumers are no better or worse off: they still pay $p_w$ and consume $q_i^D$
- The government gains revenue $R_t$

]

---
count: false
# Effects of unilateral carbon tax

.pull-left[
```{r fig3-unilateral-stage6, out.width = "100%", fig.width = 5.8, fig.height = 4.3, echo = FALSE, fig.align = "center"}
ct_plot_figure_3(stage = "welfare")
```
]

.pull-right[

From country $i$'s perspective:
- Producers lose producer surplus equal to $R_t + d$

]

---
count: false
# Effects of unilateral carbon tax

.pull-left[
```{r fig3-unilateral-stage7, out.width = "100%", fig.width = 5.8, fig.height = 4.3, echo = FALSE, fig.align = "center"}
ct_plot_figure_3(stage = "welfare")
```
]

.pull-right[

Why do producers lose $R_t + d$?
- They receive a lower net-of-tax price on the units they still sell
- They also stop selling the units between $q_{it}^S$ and $q_i^S$

]

---
count: false
# Effects of unilateral carbon tax

.pull-left[
```{r fig3-unilateral-stage8, out.width = "100%", fig.width = 5.8, fig.height = 4.3, echo = FALSE, fig.align = "center"}
ct_plot_figure_3(stage = "welfare")
```
]

.pull-right[

Why is $d$ the country's deadweight loss?
- $R_t$ is only a transfer from producers to the government, so it stays inside country $i$
- Net national effect: $+R_t - (R_t + d) = -d$

]

---
count: false
# Effects of unilateral carbon tax

.pull-left[
```{r fig3-unilateral-stage9, out.width = "100%", fig.width = 5.8, fig.height = 4.3, echo = FALSE, fig.align = "center"}
ct_plot_figure_3(stage = "welfare")
```
]

.pull-right[

Why does leakage make $d$ the relevant national loss?
- The triangle $d$ is the pure private loss left after netting out the transfer $R_t$
- Consumers in country $i$ are unchanged and imports replace the lost domestic output

]

---
count: false
# Effects of unilateral carbon tax

.pull-left[
```{r fig3-unilateral-stage10, out.width = "100%", fig.width = 5.8, fig.height = 4.3, echo = FALSE, fig.align = "center"}
ct_plot_figure_3(stage = "welfare")
```
]

.pull-right[

Why does leakage make $d$ the relevant national loss?
- So the tax cuts home production, but foreign production rises instead: that is leakage
- Because the environmental gain is undermined abroad, country $i$ is left mainly with the private distortion cost $d$

]

---

# Unilateral carbon tax effects

The increase in imports means .hi[foreign production increases]

--

This is .hi[carbon leakage]: emissions shift abroad rather than being reduced

--

Mechanism in words:

- Country $i$ raises its marginal production costs through the carbon tax
- Country $i$ producers supply less at any price with a higher MC
- World price is unchanged at $P_w$ so quantity demanded domestically for the good does not change
- Imports from abroad fill in the difference
- Production and emissions shift abroad

---

# Leakage undermines the tax

With identical carbon intensities and a small country implementing the tax so $P_W$ is fixed:
- Every unit of reduced domestic production = 1 unit more foreign production
- .hi[Net global emissions unchanged!]

--

The carbon tax:
- Generates tax revenue $R_t$ (rectangle)
- Creates deadweight loss $d$ (triangle) from production distortion
- Achieves .hi[zero environmental benefit]

--

Country $i$ bears costs but gets no environmental gain

---

# Leakage is a rate, not a constant

Economists often summarize leakage with the .hi[leakage rate]:

$$
\text{Leakage rate} = \frac{\Delta E_j}{-\Delta E_i}
$$

where $\Delta E_i < 0$ is the change in home emissions and $\Delta E_j > 0$ is the change in foreign emissions.

---

# Leakage is a rate, not a constant

$$
\text{Leakage rate} = \frac{\Delta E_j}{-\Delta E_i}
$$

Benchmarks:
- $0$: no foreign offset
- Between $0$ and $1$: global emissions still fall, but less than the domestic reduction
- $1$: one-for-one offset, as in the small-open-economy benchmark with identical intensities
- Greater than $1$: possible if output shifts to dirtier foreign producers

---

# What determines leakage?

Leakage is .hi[larger] when the world price barely moves:
- The taxing country is small in world markets, so its supply cut hardly changes the world price
- With little price movement, domestic demand changes little and imports replace the lost home output

---
count: false
# What determines leakage?

Leakage is .hi[larger] when foreign replacement is easy or dirty:
- Foreign supply is elastic and trade costs are low, so firms abroad can expand output quickly when home production falls
- Foreign production is more emissions-intensive than home production ($e_j > e_i$), so each unit shifted abroad offsets more of the home emissions reduction

---
count: false
# What determines leakage?

Leakage is .hi[smaller] when policy reduces demand, not just home supply:
- The taxing country is large in world markets, so its supply cut changes the world price
- With price increases, domestic quantity demanded falls
- Then fewer imports are needed to fill the gap, so more of the domestic emissions cut becomes a true global reduction

---
count: false
# What determines leakage?

Leakage is .hi[smaller] when foreign expansion is constrained or cleaner:
- If foreign producers face their own carbon price, they do not expand output as aggressively when home production falls
- If home production is dirtier than foreign production ($e_i > e_j$), shifting output abroad offsets less of the domestic abatement and can even lower global emissions

---

class: inverse, center, middle
name: tariff

# Carbon tax + carbon tariff

<html><div style='float:left'></div><hr color='#EB811B' size=1px width=796px></html>

---

# Adding a carbon tariff

The problem with the carbon tax was that consumers can easily switch to importing the dirty good

--

One potential way to address this is a .hi[carbon tariff] of rate $\tau$: a tax on imports proportional to their carbon content

--

Here we will tax imports at the same rate as domestic production: $$\tau = t = e \times d$$

---

# Adding a carbon tariff

.pull-left[
```{r fig4-tax-only-stage1, out.width = "100%", fig.width = 5.8, fig.height = 4.3, echo = FALSE, fig.align = "center"}
ct_plot_figure_4(stage = "tax_only")
```
]

.pull-right[

Start from the carbon tax alone:
- Producers net $P_W - t$
- Home output falls to $q_{it}^S$
- Consumers still pay $P_W$ and buy $q_i^D$

]

---
count: false

# Adding a carbon tariff

.pull-left[
```{r fig4-tax-only-stage2, out.width = "100%", fig.width = 5.8, fig.height = 4.3, echo = FALSE, fig.align = "center"}
ct_plot_figure_4(stage = "tax_only")
```
]

.pull-right[

Why is leakage the problem here?
- Imports replace the lost domestic output
- So home emissions fall, but foreign production rises
- The tax changes domestic supply, not consumption
- Since consumption doesn't change, there must be more imports

]

---
count: false

# Adding a carbon tariff

.pull-left[
```{r fig4-tax-tariff-stage1, out.width = "100%", fig.width = 5.8, fig.height = 4.3, echo = FALSE, fig.align = "center"}
ct_plot_figure_4(stage = "tariff")
```
]

.pull-right[

Now add a carbon tariff $\tau = t$ on imports:
- Producers can now charge $P+\tau$ so consumers pay $P_W + \tau$
- Domestic producers only net $P_W + \tau - t = P_W$ after the carbon tax payment
- The tax and tariff offset for producers so domestic supply returns to $q_i^S$

]

---
count: false

# Adding a carbon tariff

.pull-left[
```{r fig4-tax-tariff-stage2, out.width = "100%", fig.width = 5.8, fig.height = 4.3, echo = FALSE, fig.align = "center"}
ct_plot_figure_4(stage = "tariff")
```
]

.pull-right[

What changes in demand and trade?
- Consumers pay $P_W + \tau$ so domestic quantity demanded falls to $q_{i\tau}^D$
- Imports shrink from $q_i^D - q_{it}^S$ to $q_{i\tau}^D - q_i^S$

]

---
count: false

# Adding a carbon tariff

.pull-left[
```{r fig4-tax-tariff-stage2.1, out.width = "100%", fig.width = 5.8, fig.height = 4.3, echo = FALSE, fig.align = "center"}
ct_plot_figure_4(stage = "tariff")
```
]

.pull-right[

Total supply and demand effects of tax+tariff:
- No net effect on producers since their quantity is the same as unregulated
- Consumers face a higher price $P_W + \tau$, reduce consumption, and reduce imports

]

---
count: false

# Adding a carbon tariff

.pull-left[
```{r fig4-tax-tariff-stage3, out.width = "100%", fig.width = 5.8, fig.height = 4.3, echo = FALSE, fig.align = "center"}
ct_plot_figure_4(stage = "revenue")
```
]

.pull-right[

Revenue effects:
- The domestic carbon tax still raises $R_t$
- The tariff raises $R_\tau$ on remaining imports
- Both are transfers inside country $i$: tax $R_t$ paid by producers, tariff $R_\tau$ paid by consumers

]

---
count: false

# Adding a carbon tariff

.pull-left[
```{r fig4-tax-tariff-stage4, out.width = "100%", fig.width = 5.8, fig.height = 4.3, echo = FALSE, fig.align = "center"}
ct_plot_figure_4(stage = "welfare")
```
]

.pull-right[

Welfare and environment:
- $b$ is the consumption deadweight loss from the higher price
- $b+f$ is the environmental gain from lower world output and emissions
- Tax + tariff improves social welfare by $f$

]

---

# Recap

With only a tax:
- Domestic production/emissions become more expensive
- Consumers can easily switch to imports: "leakage"
- We taxed production but received no environmental benefits

Adding a carbon tariff:
- We are now making foreign production/emissions more expensive for our domestic consumers
- Consumers can no longer as easily switch to imports
- Emissions fall because they are correctly priced (good!)
- Consumption falls (bad!)

---

# The key insight

With identical carbon intensities everywhere:

--

.hi[Carbon tax + carbon tariff] = .hi[Consumption tax]

--

Why?
- Tax on domestic production raises domestic costs
- Tariff on imports raises import costs
- Both domestic and imported goods face the same carbon price
- Equivalent to taxing consumption of the carbon-intensive good

---

class: inverse, center, middle
name: tariff-alone

# Carbon tariff alone

<html><div style='float:left'></div><hr color='#EB811B' size=1px width=796px></html>

---

# What if we only use a tariff?

Some proposals call for carbon tariffs .hi[without] a domestic carbon tax

--

What does this accomplish?

---

# Carbon tariff alone

.pull-left[
```{r fig5-tariff-alone-stage1, out.width = "100%", fig.width = 5.8, fig.height = 4.3, echo = FALSE, fig.align = "center"}
ct_plot_figure_5(stage = "baseline")
```
]

.pull-right[

Start from free trade:
- Consumers and producers in country $i$ both face $P_W$
- Domestic output is $q_i^S$ and consumption is $q_i^D$
- Imports fill the gap $q_i^D - q_i^S$

]

---
count: false

# Carbon tariff alone

.pull-left[
```{r fig5-tariff-alone-stage2, out.width = "100%", fig.width = 5.8, fig.height = 4.3, echo = FALSE, fig.align = "center"}
ct_plot_figure_5(stage = "tariff")
```
]

.pull-right[

Now impose a carbon tariff $\tau$ with no domestic carbon tax:
- Consumers pay $P_W + \tau$
- Domestic producers also receive $P_W + \tau$
- So consumption falls and home output rises

]

---
count: false

# Carbon tariff alone

.pull-left[
```{r fig5-tariff-alone-stage3, out.width = "100%", fig.width = 5.8, fig.height = 4.3, echo = FALSE, fig.align = "center"}
ct_plot_figure_5(stage = "tariff")
```
]

.pull-right[

Why is this different from tax + tariff?
- The tariff alone taxes consumption but also protects domestic producers
- Output rises to $q_{i\tau}^S$ and demand falls to $q_{i\tau}^D$
- Imports shrink because both margins move

]

---
count: false

# Carbon tariff alone

.pull-left[
```{r fig5-tariff-alone-stage4, out.width = "100%", fig.width = 5.8, fig.height = 4.3, echo = FALSE, fig.align = "center"}
ct_plot_figure_5(stage = "revenue")
```
]

.pull-right[

- The government collects tariff revenue $R_\tau$ on remaining imports
- The tariff pushes some production away from cheaper foreign suppliers

]

---
count: false

# Carbon tariff alone

.pull-left[
```{r fig5-tariff-alone-stage5, out.width = "100%", fig.width = 5.8, fig.height = 4.3, echo = FALSE, fig.align = "center"}
ct_plot_figure_5(stage = "welfare")
```
]

.pull-right[

Welfare and environment:
- $a$ is the production deadweight loss from replacing cheaper imports with costlier home output
- Reduced consumption creates environmental gains of $b+f$, offsetting the consumption loss $b$

]

---
count: false

# Carbon tariff alone

.pull-left[
```{r fig5-tariff-alone-stage5.1, out.width = "100%", fig.width = 5.8, fig.height = 4.3, echo = FALSE, fig.align = "center"}
ct_plot_figure_5(stage = "welfare")
```
]

.pull-right[

Welfare and environment:
- Net social effect is $(b+f)-(b+a) = f-a$, so a tariff alone mixes climate gains with a subsidy to domestic pollution

]

---

class: inverse, center, middle
name: world-tax

# World carbon tax

<html><div style='float:left'></div><hr color='#EB811B' size=1px width=796px></html>

---

# World carbon tax

<!-- FIGURE 6: World carbon tax showing efficient outcome, tariff unnecessary -->

.pull-left[
```{r fig6-world-tax, out.width = "100%", fig.width = 5.8, fig.height = 4.3, echo = FALSE, fig.align = "center"}
ct_plot_figure_6()
```
]

.pull-right[

Suppose the rest of the world already prices carbon so $P_W$ internalizes the externality *abroad*:
- $P_W$ is then where abroad MSC = abroad D
- $i$ produces at $q_i^S$, but at $q_i^S$: MSC $>$ world MB given by $P_W$ leading to DWL $a$
- If $i$ also imposes the carbon tax $t$, supply falls from $q_i^S$ to $q_{it}^S$
- Imports rise to $q_i^D-q_{it}^S$

]


---

# World carbon tax

<!-- FIGURE 6: World carbon tax showing efficient outcome, tariff unnecessary -->

.pull-left[
```{r fig6-world-tax.1, out.width = "100%", fig.width = 5.8, fig.height = 4.3, echo = FALSE, fig.align = "center"}
ct_plot_figure_6()
```
]

.pull-right[

If country $i$ also imposes the carbon tax $t$:
- Producers receive net-of-tax price $P_W - t$
- Supply falls from $q_i^S$ to $q_{it}^S$
- Imports rise to $q_i^D-q_{it}^S$ to make up the shortfall given quantity demanded does not change given world price $P_W$

]

---
count: false

# World carbon tax

.pull-left[
```{r fig6-world-tax-stage2, out.width = "100%", fig.width = 5.8, fig.height = 4.3, echo = FALSE, fig.align = "center"}
ct_plot_figure_6()
```
]

.pull-right[

Why are those extra imports not leakage?
- Foreign producers already face the same carbon price
- That foreign carbon price led to leakage *into* country $i$
- Those imports reverse leakage from the rest of the world to country $i$
- The DWL $a$ disappears

]

---
count: false

# World carbon tax

.pull-left[
```{r fig6-world-tax-stage3, out.width = "100%", fig.width = 5.8, fig.height = 4.3, echo = FALSE, fig.align = "center"}
ct_plot_figure_6()
```
]

.pull-right[

At that point the allocation is efficient:
- Every country has internalized the externality
- Global marginal social cost equals the marginal benefit of the last unit consumed
- A carbon tariff is unnecessary

]

---

class: inverse, center, middle
name: complications

# Complications

<html><div style='float:left'></div><hr color='#EB811B' size=1px width=796px></html>

---

# Real-world complications

Several factors complicate carbon tariff design:

--

1. .hi[Resource shuffling:] clean production exported, dirty production consumed domestically

--

2. .hi[Non-price policies:] regulations vs. carbon prices

--

3. .hi[Export rebates:] should exports get carbon tax rebates?

--

4. .hi[Retaliation:] trading partners may impose counter-tariffs

---

# Resource shuffling

Suppose country $j$ has both clean and dirty production

--

With a carbon tariff based on average intensity:
- Export the clean stuff to country $i$
- Consume the dirty stuff domestically

--

No change in country $j$'s total emissions!

--

This is .hi[another form of leakage]

---

# Non-price policies

Many countries use regulations rather than carbon prices:
- Fuel economy standards
- Renewable portfolio standards
- Building codes

--

How do you calculate the "equivalent carbon price"?

--

Very difficult: may require complex modeling

---

# Export rebates

Should country $i$ rebate carbon taxes on exports?

--

Arguments for:
- Maintains competitiveness
- Symmetric with import tariffs

--

Arguments against:
- Reduces domestic incentive to decarbonize
- May look like an export subsidy (WTO issues)

---

# WTO compatibility

Carbon tariffs face WTO scrutiny:

--

1. .hi[National treatment:] can't discriminate against foreign goods

--

2. .hi[Most favored nation:] can't discriminate among trading partners

--

3. .hi[GATT Article XX:] exceptions for environmental protection

--

EU CBAM designed carefully to fit within WTO rules

---

class: inverse, center, middle
name: takeaways

# Key takeaways

<html><div style='float:left'></div><hr color='#EB811B' size=1px width=796px></html>

---

# Summary

1. .hi[Carbon tariffs] aim to address leakage and competitiveness concerns

--

2. .hi[Carbon tax + tariff] is equivalent to a consumption tax
   - Reduces consumption of carbon-intensive goods
   - Can improve welfare if environmental benefit exceeds distortion costs

--

3. .hi[Carbon tariff alone] = consumption tax + production subsidy
   - May increase domestic emissions
   - Less effective than tax + tariff

---

# Summary (continued)

4. .hi[World carbon tax] is first-best
   - Carbon tariffs are second-best when global coordination fails

--

5. .hi[Different carbon intensities] create measurement challenges
   - Setting tariff = foreign intensity is efficient but looks protectionist
   - Setting tariff = domestic intensity under-prices foreign emissions

--

6. .hi[Many complications] in practice
   - Resource shuffling, non-price policies, export rebates, WTO rules

---

# The bottom line

Carbon tariffs are a .hi[complement] to carbon pricing, not a substitute

--

They work best when:
- Combined with a domestic carbon tax
- Set based on actual carbon content
- Part of broader international coordination

--

Alone, they are an imperfect tool that may create new distortions
