---
title: "Lecture 09"
subtitle: "Carbon Tariffs"
author: Ivan Rudik
date: AEM 4510
output:
  xaringan::moon_reader:
    css: ['default', 'metropolis', 'metropolis-fonts', 'my-css.css']
    # self_contained: true
    nature:
      highlightStyle: github
      highlightLines: true
      countIncrementalSlides: false
      ratio: '16:9'

<!-- Auto-rebuild: ls slides/09-slides-carbon-tariffs.Rmd | entr -c Rscript -e "rmarkdown::render('slides/09-slides-carbon-tariffs.Rmd')" -->

---
exclude: true
```{r setup}
if (!require("pacman")) install.packages("pacman")
pacman::p_load(
  tidyverse, xaringanExtra, rlang, patchwork, ragg
)
source("R/video_helpers.R")
source("R/09-carbon-tariffs-figures.R")
options(htmltools.dir.version = FALSE)
knitr::opts_hooks$set(fig.callout = function(options) {
  if (options$fig.callout) {
    options$echo <- FALSE
  }
  options
})
knitr::opts_chunk$set(
  cache = TRUE,
  cache.extra = list(
    tools::md5sum("R/09-carbon-tariffs-figures.R")
  ),
  echo = TRUE,
  fig.align = "center",
  dev = "ragg_png",
  dpi = 220,
  fig.retina = 2
)
```
```{r xaringanExtra, echo = FALSE}
xaringanExtra::use_xaringan_extra(c("panelset", "webcam"))

```
```{r echo=FALSE}
xaringanExtra::style_panelset(panel_tab_color_active = "red")
```

---

# Roadmap

1. What are carbon tariffs?
2. Basic trade theory: supply, demand, and tariffs
3. Adding pollution externalities to trade
4. Unilateral carbon taxes and leakage
5. Carbon tariffs to prevent leakage
6. Carbon tariffs alone
7. World carbon taxes
8. Different pollution intensities across countries

---

class: inverse, center, middle
name: overview

# What are carbon tariffs?

<html><div style='float:left'></div><hr color='#EB811B' size=1px width=796px></html>

---

# Carbon tariffs

Carbon tariffs are tariffs imposed on imports based on their .hi[carbon content]

--

Why would we want carbon tariffs?

--

1. .hi[Leakage:] if country $i$ imposes a carbon tax, production may shift to country $j$ with no carbon tax

--

2. .hi[Competitiveness:] domestic firms face higher costs than foreign competitors

--

Carbon tariffs aim to "level the playing field" and reduce leakage

---

# Carbon tariffs in practice

.hi[EU Carbon Border Adjustment Mechanism (CBAM):]
- Phasing in 2023-2026, full implementation 2026
- Covers cement, iron/steel, aluminum, fertilizers, electricity, hydrogen
- Importers must purchase certificates matching EU ETS price

--

.hi[UK CBAM:] announced for 2027

--

.hi[US proposals:] various bills proposed but none passed yet

---

# Understanding border carbon adjustments

```{r, echo = FALSE}
local_video("videos/09-carbon-tariffs-bca-explained.mp4")
```

---

# This lecture

We will use a simple partial equilibrium framework from Brunel and Levinson (2024) to understand:

--

1. How carbon tariffs work in theory

--

2. When they improve welfare

--

3. What complications arise in practice

--

The key insight: carbon tariffs are equivalent to a .hi[consumption tax] plus a .hi[production subsidy]

---

class: inverse, center, middle
name: trade

# Basic trade theory

<html><div style='float:left'></div><hr color='#EB811B' size=1px width=796px></html>

---

# Tariffs are taxes on imports

A tariff is a .hi[tax on imports]

--

In this lecture we study the .hi[small-country] case:
- country $i$ is too small to affect the world price
- imports are available at the exogenous world price $P_W$
- this is the benchmark that makes leakage especially transparent later on

---

# A simple trade model

Consider a small open economy (country $i$) that:
- Takes world prices as given
- Can import or export a homogeneous good

--

Key curves:
- $S$: domestic supply
- $D$: domestic demand
- $P_W$: world price
    - The social marginal benefit of the good

Domestic suppliers can't charge more than $P_W$, and won't charge less

---

# Supply and demand refresher

In this market:

--

- $D$ is .hi[marginal benefit] to consumers

--

- $S$ is .hi[marginal cost] to domestic producers

--

- If $P_W$ is below the autarky price, country $i$ imports the good

--

- The gap between domestic demand and domestic supply is imports

---

# Free trade benchmark

.pull-left[
```{r fig1-tariff, out.width = "100%", fig.width = 5.6, fig.height = 4.3, echo = FALSE, fig.align = "center"}
ct_plot_figure_1(stage = "free_trade")
```
]

.pull-right[

With free trade at price $P_W$:
- Domestic production is $q_i^S$
- Domestic consumption is $q_i^D$
- Imports are $q_i^D - q_i^S$

]

---
count: false
# The tariff wedge

.pull-left[
```{r fig1-tariff-stage2, out.width = "100%", fig.width = 5.6, fig.height = 4.3, echo = FALSE, fig.align = "center"}
ct_plot_figure_1(stage = "price")
```
]

.pull-right[

Suppose we implement an import tariff $\tau$:
- The domestic price rises from $P_W$ to $P_W + \tau$
- Imported units now enter at the tariff-inclusive price
- Domestic suppliers can charge that same higher price

]

---
count: false
# Quantity effects of a tariff

.pull-left[
```{r fig1-tariff-stage2.1, out.width = "100%", fig.width = 5.6, fig.height = 4.3, echo = FALSE, fig.align = "center"}
ct_plot_figure_1(stage = "quantities")
```
]

.pull-right[

A tariff $\tau$ raises the domestic price to $P_W + \tau$:
- Domestic production rises to $q_{i\tau}^S$
- Domestic consumption falls to $q_{i\tau}^D$
- Imports shrink to $q_{i\tau}^D - q_{i\tau}^S$

]

---
count: false
# Consumer surplus refresher

.pull-left[
```{r fig1-consumer-surplus, out.width = "100%", fig.width = 5.6, fig.height = 4.3, echo = FALSE, fig.align = "center"}
ct_plot_surplus_refresher(type = "consumer")
```
]

.pull-right[

Consumer surplus is the area below demand and above price.

- Blue rectangle: consumers pay more on the units they still buy
- Orange triangle: consumers stop buying units they still valued above $P_W$

]

---
count: false
# Producer surplus refresher

.pull-left[
```{r fig1-producer-surplus, out.width = "100%", fig.width = 5.6, fig.height = 4.3, echo = FALSE, fig.align = "center"}
ct_plot_surplus_refresher(type = "producer")
```
]

.pull-right[

Producer surplus is the area above supply and below price.

- Blue rectangle: producers receive a higher price on their original output
- Orange triangle: producers gain extra surplus when they expand output to $q_{i\tau}^S$

]

---
count: false
# Welfare accounting under a tariff

.pull-left[
```{r fig1-tariff-stage3, out.width = "100%", fig.width = 5.6, fig.height = 4.3, echo = FALSE, fig.align = "center"}
ct_plot_figure_1(stage = "welfare_components")
```
]

.pull-right[

- $a$: producer gain on the original domestic output
- $b$: producer gain on the extra domestic output
- $c$: production deadweight loss
- $d$: tariff revenue on remaining imports
- $e$: consumption deadweight loss

]

---
count: false
# Consumer loss and redistribution

.pull-left[
```{r fig1-tariff-stage3.05, out.width = "100%", fig.width = 5.6, fig.height = 4.3, echo = FALSE, fig.align = "center"}
ct_plot_figure_1(stage = "welfare_components")
```
]

.pull-right[

- $-\Delta CS = a + b + c + d + e$
- The whole shaded area is the .hi[consumer surplus loss]
- But $a+b$ reappear as part of .hi[producer surplus gain]
- And $d$ reappears as .hi[tariff revenue]

]

---
count: false
# Tariff welfare accounting

.pull-left[
```{r fig1-tariff-stage3.1, out.width = "100%", fig.width = 5.6, fig.height = 4.3, echo = FALSE, fig.align = "center"}
ct_plot_figure_1(stage = "welfare_dwl")
```
]

.pull-right[

- $\Delta PS = a + b$
- $\Delta CS = -(a + b + c + d + e)$
- $R = d$
- $\Delta W_i = \Delta PS + R - \Delta CS$
- So for a .hi[small country], $\Delta W_i = -c - e$
- The only net welfare losses are the two deadweight-loss regions $c$ and $e$

]

---
count: false
# Production distortion

.pull-left[
```{r fig1-tariff-stage4, out.width = "100%", fig.width = 5.6, fig.height = 4.3, echo = FALSE, fig.align = "center"}
ct_plot_figure_1(stage = "production_distortion")
```
]

.pull-right[

Triangle $c$ is a production distortion:
- The tariff shifts production from lower-cost foreign supply to higher-cost domestic production
- For those extra domestic units, MC ($S[i]$) exceeds MB ($P_W$)

]

---
count: false
# Consumption distortion

.pull-left[
```{r fig1-tariff-stage5, out.width = "100%", fig.width = 5.6, fig.height = 4.3, echo = FALSE, fig.align = "center"}
ct_plot_figure_1(stage = "consumption_distortion")
```
]

.pull-right[

Triangle $e$ is a consumption distortion:
- The tariff eliminates units consumers value above the world price but below the tariff-inclusive price
- For those lost units, MB ($D[i]$) exceeds MC ($P_W$)

]

---

# Why start with ordinary tariffs?

In a .hi[small open economy], an ordinary tariff is a private welfare loss: it transfers surplus around inside the country but leaves deadweight loss $c+e$

--

- Consumer surplus falls by $a+b+c+d+e$
- Producer surplus rises by $a+b$
- Government revenue is $d$, so the private net loss is the deadweight loss $c+e$

--

Carbon tariffs are only potentially useful because there is .hi[another distortion] missing from this diagram: production creates pollution

--

So next we add the environmental externality to this standard tariff benchmark

---

class: inverse, center, middle
name: pollution

# Adding pollution

<html><div style='float:left'></div><hr color='#EB811B' size=1px width=796px></html>

---

# Production creates pollution

Now suppose production of this good creates pollution

--

Let $e$ = emissions per unit of output (same everywhere for now)

--

Let $d$ = marginal damage from emissions (global pollutant)

--

The .hi[social marginal cost (SMC)] of production is: $SMC = S + e \times d$

--

SMC lies above the private supply curve by the external cost $e \times d$

---

# Free trade with pollution

<!-- FIGURE 2a: Country i with SMC curve above S, showing pollution externality -->

.pull-left[
```{r fig2a-pollution, out.width = "100%", fig.width = 5.4, fig.height = 4.3, echo = FALSE, fig.align = "center"}
ct_plot_figure_2_country(stage = "baseline")
```
]

.pull-right[

Free trade output is pinned down by domestic demand and supply at the world price

]

---
count: false
# Free trade with pollution

.pull-left[
```{r fig2a-pollution-stage2, out.width = "100%", fig.width = 5.4, fig.height = 4.3, echo = FALSE, fig.align = "center"}
ct_plot_figure_2_country(stage = "msc")
```
]

.pull-right[

Country $i$'s SMC curve lies above its private supply curve

The vertical gap is the external cost: $e \times d$

]

---
count: false
# Free trade with pollution

.pull-left[
```{r fig2a-pollution-stage3, out.width = "100%", fig.width = 5.4, fig.height = 4.3, echo = FALSE, fig.align = "center"}
ct_plot_figure_2_country(stage = "damage")
```
]

.pull-right[

The green region is the domestic deadweight loss from producing beyond the efficient output level

]

---
count: false
# Free trade with pollution

.pull-left[
```{r fig2a-pollution-stage4, out.width = "100%", fig.width = 5.4, fig.height = 4.3, echo = FALSE, fig.align = "center"}
ct_plot_figure_2_country(stage = "damage")
```
]

.pull-right[

At the world price $P_W$, marginal benefit is pinned down by world demand, while marginal social cost is given by $SMC$

That triangle is the set of units for which marginal benefit exceeds private supply cost but is below marginal social cost

]

---

# Global perspective

<!-- FIGURE 2b: World supply and demand with SMC -->

.pull-left[
```{r fig2b-world, out.width = "100%", fig.width = 5.4, fig.height = 4.3, echo = FALSE, fig.align = "center"}
ct_plot_figure_2_world()
```
]

.pull-right[

From a global perspective:
- Free-trade output is where $D = S$
- Efficient output is where $D = SMC$
- The green region is the global welfare gain from cutting output from $Q_W$ to $Q_W^*$

]

---

# Country and global perspectives

```{r fig2-comparison, out.width = "100%", fig.width = 11, fig.height = 4.4, echo = FALSE, fig.align = "center"}
patchwork::wrap_plots(
  ct_plot_figure_2_country(stage = "damage"),
  ct_plot_figure_2_world(),
  ncol = 2,
  widths = c(1, 1.1)
)
```

Global DWL on the right is the sum of each country's DWL on the left

---

# The externality problem

With a global pollutant like $CO_2$:

--

- Damages occur regardless of where emissions happen
- No country has incentive to fully internalize the externality
- Free trade leads to .hi[too much production] globally

--

A global carbon tax $= e \times d$ would achieve efficiency

--

But what if only .hi[one country] acts?

---

class: inverse, center, middle
name: unilateral

# Unilateral carbon tax

<html><div style='float:left'></div><hr color='#EB811B' size=1px width=796px></html>

---

# Country $i$ imposes a carbon tax

Suppose country $i$ imposes a carbon tax $t = e \times d$ on domestic production

--

This shifts country $i$'s supply curve up by $t$

--

But the .hi[world price doesn't change] (small open economy)

--

So the carbon tax reduces the net price domestic producers receive from $P_W$ to $P_W - t$

---

# Effects of unilateral carbon tax

<!-- FIGURE 3: Unilateral carbon tax showing tax revenue Rt, deadweight loss d, and leakage -->

.pull-left[
```{r fig3-unilateral, out.width = "100%", fig.width = 5.8, fig.height = 4.3, echo = FALSE, fig.align = "center"}
ct_plot_figure_3(stage = "baseline")
```
]

.pull-right[

Before the tax, the domestic market is at the free-trade benchmark with price $P_W$

- You can think of a tax $t = e \times d$ as shifting $S[i]$ up to $SMC[i]$, so firms produce until $S[i] + t = P_W$
- Equivalently, firms keep the same supply curve $S[i]$ but face the net-of-tax price $P_W - t$

]

---
count: false
# Effects of unilateral carbon tax

.pull-left[
```{r fig3-unilateral-stage2, out.width = "100%", fig.width = 5.8, fig.height = 4.3, echo = FALSE, fig.align = "center"}
ct_plot_figure_3(stage = "tax")
```
]

.pull-right[

After the tax:
- Consumers still face the world price $P_W$
- But with a tax $t$, producers receive the net-of-tax price $P_W - t$
- Domestic output falls to $q_{it}^S$

]

---
count: false
# Effects of unilateral carbon tax

.pull-left[
```{r fig3-unilateral-stage3, out.width = "100%", fig.width = 5.8, fig.height = 4.3, echo = FALSE, fig.align = "center"}
ct_plot_figure_3(stage = "tax_delta_imports")
```
]

.pull-right[

.hi[Leakage] comes next:
- Consumers in country $i$ still demand $q_i^D$ at price $P_W$
- Imports rise enough to offset the fall in domestic output
- Foreign production replaces home production: that is leakage
- .hi[Global] quantity produced (and emissions) does not change

]

---
count: false
# Effects of unilateral carbon tax

.pull-left[
```{r fig3-unilateral-stage4, out.width = "100%", fig.width = 5.8, fig.height = 4.3, echo = FALSE, fig.align = "center"}
ct_plot_figure_3(stage = "welfare")
```
]

.pull-right[

From country $i$'s perspective:
- Consumer surplus does not change: they still pay $P_W$ and consume $q_i^D$
- The government gains revenue $R_t$

]

---
count: false
# Effects of unilateral carbon tax

.pull-left[
```{r fig3-unilateral-stage6, out.width = "100%", fig.width = 5.8, fig.height = 4.3, echo = FALSE, fig.align = "center"}
ct_plot_figure_3(stage = "welfare")
```
]

.pull-right[

From country $i$'s perspective:
- Producers lose producer surplus equal to $R_t + d$

]

---
count: false
# Effects of unilateral carbon tax

.pull-left[
```{r fig3-unilateral-stage7, out.width = "100%", fig.width = 5.8, fig.height = 4.3, echo = FALSE, fig.align = "center"}
ct_plot_figure_3(stage = "welfare")
```
]

.pull-right[

Why do producers lose $R_t + d$?
- Blue rectangle: they receive a lower net-of-tax price on the units they still sell
- Orange triangle: they also stop selling the units between $q_{it}^S$ and $q_i^S$ for which their PMC $< P_W$

]

---
count: false
# Effects of unilateral carbon tax

.pull-left[
```{r fig3-unilateral-stage8, out.width = "100%", fig.width = 5.8, fig.height = 4.3, echo = FALSE, fig.align = "center"}
ct_plot_figure_3(stage = "welfare")
```
]

.pull-right[

Why is $d$ the country's deadweight loss?
- $R_t$ is only a transfer from producers to the government, so it stays inside country $i$
- Net national effect: $+R_t - (R_t + d) = -d$

]

---
count: false
# Effects of unilateral carbon tax

.pull-left[
```{r fig3-unilateral-stage9, out.width = "100%", fig.width = 5.8, fig.height = 4.3, echo = FALSE, fig.align = "center"}
ct_plot_figure_3(stage = "welfare")
```
]

.pull-right[

Why does leakage make $d$ the relevant national loss?
- The triangle $d$ is the pure private loss left after netting out the transfer $R_t$
- Consumers in country $i$ are unaffected and imports replace the lost domestic output/emissions

]

---
count: false
# Effects of unilateral carbon tax

.pull-left[
```{r fig3-unilateral-stage10, out.width = "100%", fig.width = 5.8, fig.height = 4.3, echo = FALSE, fig.align = "center"}
ct_plot_figure_3(stage = "welfare")
```
]

.pull-right[

Why does leakage make $d$ the relevant national loss?
- The tax reduces home production, but foreign production rises 1-for-1: that is leakage
- The environmental gain is offset abroad, so country $i$ is left with the private DWL $d$

]

---

# Unilateral carbon tax effects

The increase in imports means .hi[foreign production increases]

--

This is .hi[carbon leakage]: emissions shift abroad rather than being reduced

--

- Country $i$ raises its marginal production costs through the carbon tax

--

- Country $i$ producers supply less at any price with a higher MC

--

- World price is unchanged at $P_W$ so quantity demanded domestically for the good does not change

--

- Imports from abroad fill in the difference

--

- Production and emissions increase abroad

---

# Leakage undermines the tax

With identical carbon intensities and a small country implementing the tax (so $P_W$ is fixed):
- 1 unit fewer domestic production = 1 unit more foreign production
- .hi[Net global emissions unchanged!]

--

The carbon tax:
- Generates tax revenue $R_t$ (rectangle)
- Creates deadweight loss $d$ (triangle) from production distortion
- Achieves .hi[zero environmental benefit]

--

Country $i$ bears costs but gets no environmental gain

---

# Leakage is a rate, not a constant

Economists often summarize leakage with the .hi[leakage rate]:

$$
\text{Leakage rate} = \frac{\Delta E_j}{-\Delta E_i}
$$

where $\Delta E_i < 0$ is the change in home emissions and $\Delta E_j > 0$ is the change in foreign emissions

---

# Leakage is a rate, not a constant

$$
\text{Leakage rate} = \frac{\Delta E_j}{-\Delta E_i}
$$

Benchmarks:
- $0$: no foreign offset
- Between $0$ and $1$: global emissions still fall, but less than the domestic reduction
- $1$: one-for-one offset, as in this small-open-economy example with identical intensities
- Greater than $1$: possible if output shifts to dirtier foreign producers

---

# What determines leakage?

Leakage is .hi[larger] when the world price moves less in response to the tax:

--

- The taxing country is small in world markets, so its supply cut hardly changes the world price

--

- With little price movement, domestic demand doesn't change and imports replace the lost domestic output

---
count: false
# What determines leakage?

Leakage is .hi[larger] when foreign replacement is elastic or dirty:

--

- Foreign supply is elastic and trade costs are low, so firms abroad can expand output quickly when home production falls

--

- Foreign production is more emissions-intensive than home production ($e_j > e_i$), so each unit shifted abroad offsets more of the home emissions reduction

---
count: false
# What determines leakage?

Leakage is .hi[smaller] when policy reduces quantity demanded by increasing the world price:

--

- The taxing country is large in world markets, so its supply cut changes the world price

--

- With price increases, domestic quantity demanded falls

--

- Then fewer imports are needed to fill the gap, so there is less of an emissions increase abroad, and greater global emissions reductions

---
count: false
# What determines leakage?

Leakage is .hi[smaller] when foreign expansion is constrained or cleaner:

--

- If foreign producers supply is inelastic, firms abroad are limited in how much they can increase production

--

- If home production is dirtier than foreign production ($e_i > e_j$), shifting output abroad offsets less of the domestic abatement and can even lower global emissions

---

class: inverse, center, middle
name: tariff

# Carbon tax + carbon tariff

<html><div style='float:left'></div><hr color='#EB811B' size=1px width=796px></html>

---

# Adding a carbon tariff

The problem with the carbon tax was that consumers can easily switch to importing the dirty good from another country not subject to the carbon tax

--

One potential way to address this is a .hi[carbon tariff] of rate $\tau$: a tax on imports proportional to their carbon content

--

Here we will tax imports at the same rate as domestic production: $$\tau = t = e \times d$$

---
count: false

# Adding a carbon tariff

.hi[Key intuition:]

A carbon tax reduces the net price producers receive by $t$

--

A carbon tariff raises the price producers receive by $\tau = t$, so producers get $P_W - t + \tau = P_W$, the price without a tax or tariff

--

A carbon tariff raises the price consumers pay by $\tau = t$, so consumers pay $P_W + \tau$

--

So tax + tariff works like a consumption tax in the end

---

# Adding a carbon tariff

.pull-left[
```{r fig4-tax-only-stage1, out.width = "100%", fig.width = 5.8, fig.height = 4.3, echo = FALSE, fig.align = "center"}
ct_plot_figure_4(stage = "tax_only")
```
]

.pull-right[

Start from the carbon tax alone:
- Producers net $P_W - t$
- Home output is $q_{it}^S$
- Consumers still pay $P_W$ and buy $q_i^D$
- Country $i$ imports $q_i^D - q_{it}^S$

]

---
count: false

# Adding a carbon tariff

.pull-left[
```{r fig4-tax-only-stage2, out.width = "100%", fig.width = 5.8, fig.height = 4.3, echo = FALSE, fig.align = "center"}
ct_plot_figure_4(stage = "tax_only")
```
]

.pull-right[

Why is leakage the problem here?
- Imports replace the lost domestic output
- So home emissions fall, but foreign production rises
- The tax changes domestic supply, not consumption
- Since consumption doesn't change, there must be more imports

]

---
count: false

# Adding a carbon tariff

.pull-left[
```{r fig4-tax-tariff-stage2, out.width = "100%", fig.width = 5.8, fig.height = 4.3, echo = FALSE, fig.align = "center"}
ct_plot_figure_4(stage = "tariff")
```
] 

.pull-right[

Now add a carbon tariff $\tau = t$ on imports:
- Consumers pay $P_W + \tau$
- Domestic producers net $P_W + \tau - t = P_W$
- So domestic supply returns to $q_i^S$ 

]

---
count: false

# Adding a carbon tariff

.pull-left[
```{r fig4-tax-tariff-stage2.1, out.width = "100%", fig.width = 5.8, fig.height = 4.3, echo = FALSE, fig.align = "center"}
ct_plot_figure_4(stage = "tariff")
```
]

.pull-right[

Total supply and demand effects of tax+tariff:
- No net effect on producers since their quantity is the same as unregulated
- Consumers face a higher price $P_W + \tau$, reduce consumption, and reduce imports: .hi[global production and emissions fall]

]

---
count: false

# Adding a carbon tariff

.pull-left[
```{r fig4-tax-tariff-stage3, out.width = "100%", fig.width = 5.8, fig.height = 4.3, echo = FALSE, fig.align = "center"}
ct_plot_figure_4(stage = "revenue")
```
]

.pull-right[

Revenue effects:
- The domestic carbon tax still raises $R_t$
- The tariff raises $R_\tau$ on remaining imports
- Both are transfers inside country $i$: tax $R_t$ paid by producers, tariff $R_\tau$ paid by consumers

]

---
count: false

# Adding a carbon tariff

.pull-left[
```{r fig4-tax-tariff-stage4, out.width = "100%", fig.width = 5.8, fig.height = 4.3, echo = FALSE, fig.align = "center"}
ct_plot_figure_4(stage = "welfare")
```
]

.pull-right[

Welfare and environment:
- $b$ is the consumption deadweight loss from the higher price
- Reduced consumption where the private benefit $D[i]$ was greater than the PMC of production $P_W$

]

---
count: false

# Adding a carbon tariff

.pull-left[
```{r fig4-tax-tariff-stage4.1, out.width = "100%", fig.width = 5.8, fig.height = 4.3, echo = FALSE, fig.align = "center"}
ct_plot_figure_4(stage = "welfare")
```
]

.pull-right[

Welfare and environment:
- $b+f$ is the environmental gain from lower world output and emissions of $q_i^D - q_{i\tau}^D$
- Tax + tariff improves social welfare by $f$

]

---

# Recap

With only a tax:
- Domestic production/emissions become more expensive
- Consumers can easily switch to imports: "leakage"
- We taxed production but received no environmental benefits

Adding a carbon tariff:
- We are now making foreign production/emissions more expensive for our domestic consumers
- Consumers can no longer as easily switch to imports
- Emissions fall because they are correctly priced (good!)
- Consumption falls (bad!)

---

# The key insight

With identical carbon intensities everywhere:

--

.hi[Carbon tax + carbon tariff] = .hi[Consumption tax]

--

Why?
- Tax on domestic production raises domestic costs
- Tariff on imports raises import costs
- Both domestic and imported goods face the same carbon price
- Equivalent to taxing consumption of the carbon-intensive good

---

```{r ct-numeric-example-data, echo = FALSE, include = FALSE}
numeric_example = ct_numeric_example_outcomes()
numeric_example_parameters = ct_numeric_example_parameters()
numeric_free_trade = numeric_example |> dplyr::filter(.data$regime == "free_trade")
numeric_tax_tariff = numeric_example |> dplyr::filter(.data$regime == "tax_tariff")
numeric_consumption_tax = numeric_example |> dplyr::filter(.data$regime == "consumption_tax")
numeric_tariff_alone = numeric_example |> dplyr::filter(.data$regime == "tariff_alone")
numeric_consumption_tax_prod_subsidy = numeric_example |>
  dplyr::filter(.data$regime == "consumption_tax_prod_subsidy")
```

# Numerical example: setup

Use the same simple market for both equivalence results:

$$Q_D = 100 - P,\qquad Q_S = P - 20,\qquad P_W = 40$$

--
Definitions:
- $Q_D$ = domestic quantity demanded
- $Q_S$ = domestic quantity supplied
- $M = Q_D - Q_S$ = imports
- $P_W$ = world price

---
count: false

# Numerical example: setup

Carbon and policy terms:
- $e_H$ and $e_F$ = home and foreign emissions per unit
- $d$ = damage per unit of emissions
- $t$ = domestic carbon tax on home production
- $\tau$ = carbon tariff on imports
$$e_H = e_F = 1,\qquad d = 10,\qquad t = \tau = 10$$

---
count: false

# Numerical example: setup

Start from the free-trade benchmark:

--

Step 1: under free trade, the domestic price equals the world price
$$P = P_W = 40$$

--

Step 2: plug $P_W$ into demand and supply
$$Q_D = 100 - 40 = `r numeric_free_trade$domestic_consumption`$$
$$Q_S = 40 - 20 = `r numeric_free_trade$domestic_production`$$
$$M = Q_D - Q_S = `r numeric_free_trade$domestic_consumption` - `r numeric_free_trade$domestic_production` = `r numeric_free_trade$imports`$$

---

# Numerical example: tax + tariff

Under a domestic carbon tax and equal tariff, work in three steps:

--

Step 1: solve for prices

Here $P_c$ is the price paid by consumers and $P_p$ is the net price received by domestic producers
$$P_c = P_W + \tau = 40 + 10 = `r numeric_tax_tariff$consumer_price`$$
$$P_p = P_c - t = `r numeric_tax_tariff$consumer_price` - 10 = `r numeric_tax_tariff$producer_price`$$

---
count: false

# Numerical example: tax + tariff

Step 2: plug those prices into demand and supply
$$Q_D = 100 - P_c = 100 - `r numeric_tax_tariff$consumer_price` = `r numeric_tax_tariff$domestic_consumption`$$
$$Q_S = P_p - 20 = `r numeric_tax_tariff$producer_price` - 20 = `r numeric_tax_tariff$domestic_production`$$
$$M = Q_D - Q_S = `r numeric_tax_tariff$domestic_consumption` - `r numeric_tax_tariff$domestic_production` = `r numeric_tax_tariff$imports`$$

---
count: false

# Numerical example: tax + tariff

Step 3: compute government revenue

Here $R_t$ is domestic carbon-tax revenue and $R_\tau$ is tariff revenue
$$R_t = t \times Q_S = 10 \times `r numeric_tax_tariff$domestic_production` = `r numeric_tax_tariff$carbon_tax_revenue`$$
$$R_\tau = \tau \times M = 10 \times `r numeric_tax_tariff$imports` = `r numeric_tax_tariff$tariff_revenue`$$
$$R = R_t + R_\tau = `r numeric_tax_tariff$carbon_tax_revenue` + `r numeric_tax_tariff$tariff_revenue` = `r numeric_tax_tariff$net_revenue`$$

---

# Numerical example: equivalent consumption tax

Now replace tax + tariff with a pure consumption tax of `r numeric_example_parameters$consumption_tax`:

--

Step 1: solve for prices

Consumers pay the tax, so $P_c$ rises, but producers still receive the world price
$$P_c = P_W + 10 = 40 + 10 = `r numeric_consumption_tax$consumer_price`$$
$$P_p = P_W = `r numeric_consumption_tax$producer_price`$$

---

# Numerical example: equivalent consumption tax

Step 2: plug those prices into demand and supply
$$Q_D = 100 - P_c = 100 - `r numeric_consumption_tax$consumer_price` = `r numeric_consumption_tax$domestic_consumption`$$
$$Q_S = P_p - 20 = `r numeric_consumption_tax$producer_price` - 20 = `r numeric_consumption_tax$domestic_production`$$
$$M = Q_D - Q_S = `r numeric_consumption_tax$domestic_consumption` - `r numeric_consumption_tax$domestic_production` = `r numeric_consumption_tax$imports`$$

---

# Numerical example: equivalent consumption tax

Step 3: compute government revenue

$$R_c = 10 \times Q_D = 10 \times `r numeric_consumption_tax$domestic_consumption` = `r numeric_consumption_tax$consumption_tax_revenue`$$

--

So tax + tariff and a consumption tax generate the same outcome:
- same consumer price: $P_c = `r numeric_tax_tariff$consumer_price`$
- same producer price: $P_p = `r numeric_tax_tariff$producer_price`$
- same domestic demand: $Q_D = `r numeric_tax_tariff$domestic_consumption`$
- same domestic supply: $Q_S = `r numeric_tax_tariff$domestic_production`$
- same imports: $M = `r numeric_tax_tariff$imports`$
- same total revenue: `r numeric_tax_tariff$net_revenue`

---

class: inverse, center, middle
name: tariff-alone

# Carbon tariff alone

<html><div style='float:left'></div><hr color='#EB811B' size=1px width=796px></html>

---

# What if we only use a tariff?

Some proposals call for carbon tariffs .hi[without] a domestic carbon tax

--

What does this accomplish?

---
count: false

# What if we only use a tariff?

.hi[Key intuition:]

A carbon tariff raises the price producers receive by $\tau$, which is the same thing as a production subsidy of $\tau$

--

A carbon tariff raises the price consumers pay by $\tau$, which is the same thing as a consumption tax of $\tau$

--

So tariff alone works like a consumption tax plus a production subsidy in the end

---

# Carbon tariff alone in practice

```{r, echo = FALSE}
local_video("videos/09-carbon-tariffs-tariff-alone-practice.mp4")
```

---

# Carbon tariff alone

.pull-left[
```{r fig5-tariff-alone-stage1, out.width = "100%", fig.width = 5.8, fig.height = 4.3, echo = FALSE, fig.align = "center"}
ct_plot_figure_5(stage = "baseline")
```
]

.pull-right[

Start from free trade:
- Consumers and producers in country $i$ both face $P_W$
- Domestic output is $q_i^S$ and consumption is $q_i^D$
- Imports fill the gap $q_i^D - q_i^S$

]

---
count: false

# Carbon tariff alone

.pull-left[
```{r fig5-tariff-alone-stage2, out.width = "100%", fig.width = 5.8, fig.height = 4.3, echo = FALSE, fig.align = "center"}
ct_plot_figure_5(stage = "tariff")
```
]

.pull-right[

Now impose a carbon tariff $\tau$ with no domestic carbon tax:
- Consumers pay $P_W + \tau$
- Domestic producers also receive $P_W + \tau$
- So consumption falls and home output rises

]

---
count: false

# Carbon tariff alone

.pull-left[
```{r fig5-tariff-alone-stage3, out.width = "100%", fig.width = 5.8, fig.height = 4.3, echo = FALSE, fig.align = "center"}
ct_plot_figure_5(stage = "tariff")
```
]

.pull-right[

Why is this different from tax + tariff?
- The tariff alone taxes consumption but also protects domestic producers
- Output rises to $q_{i\tau}^S$ and demand falls to $q_{i\tau}^D$
- Imports shrink because both margins move

]

---
count: false

# Carbon tariff alone

.pull-left[
```{r fig5-tariff-alone-stage4, out.width = "100%", fig.width = 5.8, fig.height = 4.3, echo = FALSE, fig.align = "center"}
ct_plot_figure_5(stage = "revenue")
```
]

.pull-right[

- The government collects tariff revenue $R_\tau$ on remaining imports
- The tariff pushes some production away from cheaper foreign suppliers

]

---
count: false

# Carbon tariff alone

.pull-left[
```{r fig5-tariff-alone-stage5, out.width = "100%", fig.width = 5.8, fig.height = 4.3, echo = FALSE, fig.align = "center"}
ct_plot_figure_5(stage = "welfare")
```
]

.pull-right[

- $a$ is the production DWL from replacing cheaper imports with costlier home output
- $b$ is the consumption DWL from reduced consumption at the higher price
- Lower consumption has environmental benefits $b+f$: $(q_i^D - q_{i\tau}^D)\times d$

]

---
count: false

# Carbon tariff alone

.pull-left[
```{r fig5-tariff-alone-stage5.1, out.width = "100%", fig.width = 5.8, fig.height = 4.3, echo = FALSE, fig.align = "center"}
ct_plot_figure_5(stage = "welfare")
```
]

.pull-right[

Welfare and environment:
- Net social benefit is $(b+f)-(b+a) = f-a$
- Gains from real reductions in pollution
- Losses from inefficient subsidization of domestic production
- Net effect is ambiguous

]


---
count: false

# Numerical example: setup

For the tariff-alone example, use the same market structure:

$$Q_D = 100 - P,\qquad Q_S = P - 20,\qquad P_W = 40$$

--

Definitions:
- $Q_D$ = domestic quantity demanded
- $Q_S$ = domestic quantity supplied
- $M = Q_D - Q_S$ = imports
- $P_c$ = price paid by consumers
- $P_p$ = price received by domestic producers

--

Policy for this case:
$$t = 0,\qquad \tau = 10$$

---
count: false

# Numerical example: setup

Start from the free-trade benchmark:

--

Free-trade benchmark:
$$P_c = P_p = P_W = 40$$
$$Q_D = 100 - 40 = `r numeric_free_trade$domestic_consumption`$$
$$Q_S = 40 - 20 = `r numeric_free_trade$domestic_production`$$
$$M = Q_D - Q_S = `r numeric_free_trade$imports`$$

---
count: false

# Numerical example: tariff alone

With a tariff alone, work in three steps:

--

Step 1: solve for prices

Here both consumers and domestic producers face the tariff-inclusive price
$$P_c = P_W + \tau = 40 + 10 = `r numeric_tariff_alone$consumer_price`$$
$$P_p = P_c = `r numeric_tariff_alone$producer_price`$$

---
count: false

# Numerical example: tariff alone

Step 2: plug those prices into demand and supply
$$Q_D = 100 - P_c = 100 - `r numeric_tariff_alone$consumer_price` = `r numeric_tariff_alone$domestic_consumption`$$
$$Q_S = P_p - 20 = `r numeric_tariff_alone$producer_price` - 20 = `r numeric_tariff_alone$domestic_production`$$
$$M = Q_D - Q_S = `r numeric_tariff_alone$domestic_consumption` - `r numeric_tariff_alone$domestic_production` = `r numeric_tariff_alone$imports`$$

---
count: false

# Numerical example: tariff alone

Step 3: compute tariff revenue
$$R_\tau = \tau \times M = 10 \times `r numeric_tariff_alone$imports` = `r numeric_tariff_alone$tariff_revenue`$$

---
count: false

# Numerical example: equivalent tax + subsidy

Now replace the tariff with a consumption tax of `r numeric_example_parameters$consumption_tax` and a production subsidy of `r numeric_example_parameters$production_subsidy`:

--

Step 1: solve for prices

Consumers pay the consumption tax, and producers receive the world price plus the subsidy
$$P_c = P_W + 10 = 40 + 10 = `r numeric_consumption_tax_prod_subsidy$consumer_price`$$
$$P_p = P_W + 10 = 40 + 10 = `r numeric_consumption_tax_prod_subsidy$producer_price`$$

---
count: false

# Numerical example: equivalent tax + subsidy

Step 2: plug those prices into demand and supply
$$Q_D = 100 - P_c = 100 - `r numeric_consumption_tax_prod_subsidy$consumer_price` = `r numeric_consumption_tax_prod_subsidy$domestic_consumption`$$
$$Q_S = P_p - 20 = `r numeric_consumption_tax_prod_subsidy$producer_price` - 20 = `r numeric_consumption_tax_prod_subsidy$domestic_production`$$
$$M = Q_D - Q_S = `r numeric_consumption_tax_prod_subsidy$domestic_consumption` - `r numeric_consumption_tax_prod_subsidy$domestic_production` = `r numeric_consumption_tax_prod_subsidy$imports`$$

---
count: false

# Numerical example: equivalent tax + subsidy

Step 3: compute the government budget

Here $R_c$ is consumption-tax revenue
$$R_c = 10 \times Q_D = 10 \times `r numeric_consumption_tax_prod_subsidy$domestic_consumption` = `r numeric_consumption_tax_prod_subsidy$consumption_tax_revenue`$$
$$\text{Subsidy cost} = 10 \times Q_S = 10 \times `r numeric_consumption_tax_prod_subsidy$domestic_production` = `r numeric_consumption_tax_prod_subsidy$production_subsidy_cost`$$
$$R^{net} = R_c - \text{Subsidy cost} = `r numeric_consumption_tax_prod_subsidy$consumption_tax_revenue` - `r numeric_consumption_tax_prod_subsidy$production_subsidy_cost` = `r numeric_consumption_tax_prod_subsidy$net_revenue`$$

---
count: false

# Numerical example: equivalent tax + subsidy

So tariff alone and a consumption tax plus production subsidy generate the same outcome:
- same consumer price: $P_c = `r numeric_tariff_alone$consumer_price`$
- same producer price: $P_p = `r numeric_tariff_alone$producer_price`$
- same domestic demand: $Q_D = `r numeric_tariff_alone$domestic_consumption`$
- same domestic supply: $Q_S = `r numeric_tariff_alone$domestic_production`$
- same imports: $M = `r numeric_tariff_alone$imports`$
- same net government revenue: `r numeric_tariff_alone$net_revenue`

---

class: inverse, center, middle
name: world-tax

# World carbon tax

<html><div style='float:left'></div><hr color='#EB811B' size=1px width=796px></html>

---

# World carbon tax

<!-- FIGURE 6: World carbon tax showing efficient outcome, tariff unnecessary -->

.pull-left[
```{r fig6-world-tax, out.width = "100%", fig.width = 5.8, fig.height = 4.3, echo = FALSE, fig.align = "center"}
ct_plot_figure_6(stage = "baseline")
```
]

.pull-right[

Suppose the rest of the world already prices carbon so $P_W$ reflects the carbon cost *abroad*:
- Country $i$ still produces $q_i^S$ without its own carbon tax
- The efficient output for country $i$ is $q_{it}^S$, where SMC = $P_W$

]

---

# World carbon tax

<!-- FIGURE 6: World carbon tax showing efficient outcome, tariff unnecessary -->

.pull-left[
```{r fig6-world-tax.1, out.width = "100%", fig.width = 5.8, fig.height = 4.3, echo = FALSE, fig.align = "center"}
ct_plot_figure_6(stage = "baseline")
```
]

.pull-right[

- At $q_i^S$, SMC exceeds the revenue from selling the good at $P_W$
- $q_i^S - q_{it}^S$ is leakage *into* the country from carbon taxation abroad 
- $a$ is the resulting deadweight loss from overproduction domestically

]

---

# World carbon tax

<!-- FIGURE 6: World carbon tax showing efficient outcome, tariff unnecessary -->

.pull-left[
```{r fig6-world-tax.2, out.width = "100%", fig.width = 5.8, fig.height = 4.3, echo = FALSE, fig.align = "center"}
ct_plot_figure_6(stage = "tax")
```
]

.pull-right[

If country $i$ also imposes the carbon tax $t$:
- Producers receive net-of-tax price $P_W - t$
- Supply falls from $q_i^S$ to $q_{it}^S$
- Imports rise to $q_i^D-q_{it}^S$
- Consumers still face the world price $P_W$

]

---
count: false

# World carbon tax

.pull-left[
```{r fig6-world-tax-stage2, out.width = "100%", fig.width = 5.8, fig.height = 4.3, echo = FALSE, fig.align = "center"}
ct_plot_figure_6(stage = "tax")
```
]

.pull-right[

Why are those extra imports not leakage?
- Foreign producers already face the same carbon price
- The earlier distortion was leakage from the rest of the world to country $i$
- Those imports reverse that leakage
- The DWL $a$ disappears

]

---
count: false

# World carbon tax

.pull-left[
```{r fig6-world-tax-stage3, out.width = "100%", fig.width = 5.8, fig.height = 4.3, echo = FALSE, fig.align = "center"}
ct_plot_figure_6(stage = "tax")
```
]

.pull-right[

At that point the allocation is efficient:
- Every country has internalized the externality
- Global SMC equals the marginal benefit of the last unit consumed
- Production is cost-effective across countries
- A carbon tariff is unnecessary

]

---

class: inverse, center, middle
name: complications

# Complications

<html><div style='float:left'></div><hr color='#EB811B' size=1px width=796px></html>

---

# Real-world complications

Several factors complicate carbon tariff design:

--

1. .hi[Resource shuffling:] clean production exported, dirty production consumed domestically

--

2. .hi[Non-price policies:] regulations vs. carbon prices

--

3. .hi[Export rebates:] should exports get carbon tax rebates?

--

4. .hi[Retaliation:] trading partners may impose counter-tariffs

---

# Retaliation risk

Trading partners may retaliate

```{r, echo = FALSE}
local_video("videos/09-carbon-tariffs-retaliation-trade-war.mp4")
```

---

# Resource shuffling

Suppose country $j$ has both clean and dirty production

--

With a carbon tariff based on average intensity:
- Export the clean stuff to country $i$
- Consume the dirty stuff domestically

--

No change in country $j$'s total emissions!

--

This is .hi[another form of leakage]

---

# Non-price policies

Many countries use regulations rather than carbon prices:
- Fuel economy standards
- Renewable portfolio standards
- Building codes

--

How do you calculate the "equivalent carbon price"?

--

Very difficult: may require complex modeling

---

# Export rebates

Should country $i$ rebate carbon taxes on exports?

--

Arguments for:
- Maintains competitiveness
- Symmetric with import tariffs

--

Arguments against:
- Reduces domestic incentive to decarbonize
- May look like an export subsidy (WTO issues)

---

# WTO compatibility

Carbon tariffs face WTO scrutiny:

--

1. .hi[National treatment:] can't discriminate against foreign goods

--

2. .hi[Most favored nation:] can't discriminate among trading partners

--

3. .hi[GATT Article XX:] exceptions for environmental protection

--

EU CBAM designed carefully to fit within WTO rules

---
class: inverse, center, middle
name: takeaways

# Key takeaways

<html><div style='float:left'></div><hr color='#EB811B' size=1px width=796px></html>

---

# Summary

1. .hi[Carbon tariffs] aim to address leakage and competitiveness concerns

--

2. .hi[Carbon tax + tariff] is equivalent to a consumption tax
   - Reduces consumption of carbon-intensive goods
   - Can improve welfare if environmental benefit exceeds distortion costs

--

3. .hi[Carbon tariff alone] = consumption tax + production subsidy
   - May increase domestic emissions
   - Less effective than tax + tariff

---

# Summary (continued)

4. .hi[World carbon tax] is first-best
   - Carbon tariffs are second-best when global coordination fails

--

5. .hi[Many complications] in practice
   - Resource shuffling, non-price policies, export rebates, WTO rules

---

# The bottom line

Carbon tariffs are a .hi[complement] to carbon pricing, not a substitute

--

They work best when:
- Combined with a domestic carbon tax
- Set based on actual carbon content
- Part of broader international coordination

--

Alone, they are an imperfect tool that may create new distortions
